A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-10-09 latex/main.tex

We ran fast-RIPPLES\cite{Smith2023tracking} on the Viridian UShER tree
subset to the 2,475,418 samples shared with the sc2ts ARG (STAR Methods).
At $p=3$, this resulted in a total of 4111 recombination events
of which 595 could be mapped to one of the 855 recombination events in the 
sc2ts ARG.

Figure \ref{fig:ripples_sc2ts_events}A). Analysis of the descendant 
samples showed that many Pango X lineages were not associated with
recombination events (including XBB), and we therefore ran
RIPPLES with the more lenient value of $p=3$ (the default).
This resulted in a total of 4111 recombination events
of which 595 have corresponding recombination events in the 
sc2ts ARG (Figure \ref{fig:ripples_sc2ts_events}B).
Analysis of the descending samples for each event shows that 
10 Pango of the 44 X lineages present in the dataset
% NOTE: listing them like this for sorting and manipulation purposes
(%
XA, 
XAC, 
XAE, 
XAJ,
XB, 
XBB, 
XBM, 
XC, 
XL, 
XW,%
)
are associated unequivocally with a single recombination event.

XBR, XBD 
XS appears twice

Another event suggests that 
XAF 
XAL, 
XE, 
XH, 
XJ, 
XM, 
XY, 
and 
82 BA.2 samples share a recombination event [With what breakpoint and 
parents?].

Figure \ref{fig:ripples_sc2ts_events}A). 
See Document\ref{sec:detailed_ripples} and Table~\ref{tab:ripples_x_lineages}
for detailed analysis.

@@

[Rough first draft]
This estimate for effective recombination in SARS-CoV-2 is much lower 
than previous studies have suggested.
% Bolotie: , we identified 225 anomalous SARS-CoV-2 genomes of likely
% recombinant origins out of the first 87,695 genomes to be released
In an early analysis, 225 potentially recombinant sequences out 
of 87,695 samples [WHAT DATE?]\cite{Varabyou2021rapid}.
Over the same time period, XX events are present in the sc2ts ARG,
representing a total of XX samples.
In the flagship analysis for the RIPPLES method~\cite{Turakhia2022}, 
589 unique recombination events in a tree of 1.6 million samples
up to May 2021 were identified after extensive quality control.
These nodes had a combined total of 43,104 descendants, suggesting
that approximately 2.7\% of genomes were of recombinant origin.
Over the same time period, the sc2ts ARG contains 14 recombinants
passing QC with a total of 172 descendants. The ARG contains a total
501,235 samples over this time period, giving an overall total fraction of
0.034\%. Thus, the sc2ts ARG estimates 42-fold fewer recombination
events and 79-fold descendant samples over the same time period.
The numbers are not directly comparable as the published RIPPLES 
estimate is based on GISAID data which is enriched for
artefacts~\cite{Hunt2024}, but given the observations in 
the previous section on running RIPPLES on the Viridian data
it seems likely that RIPPLES is overestimating the prevalence
of recombination in SARS-CoV-2.

@@

@@

>>>>

We ran fast-RIPPLES\cite{Smith2023tracking} on the Viridian UShER tree
subset to the 2,475,418 samples shared with the sc2ts ARG (STAR Methods).
At $p=3$, this resulted in a total of 4111 recombination events.
Most of these subtend $<10$ samples (2616 singletons), with 4 events subtending 
more than 100,000 samples.
Analysis of the descending samples for each event shows that 
10 of the 44 Pango X lineages present in the dataset
% NOTE: listing them like this for sorting and manipulation purposes
(%
XA, 
XAC, 
XAE, 
XAJ,
XB, 
XBB, 
XBM, 
XC, 
XL, 
XW%
)
are associated unequivocally with a single recombination event.
7 lineages 
(%
XAL,
XAM,
XBD, 
XBR, 
XE,
XS,
XY%
)
are associated with two events. Finally, 
one event subtending 1321 samples suggests that 
XAF,
XAL, 
XE, 
XH, 
XJ, 
XM, 
XY, 
and 
82 BA.2 samples share a recombination event [With what breakpoint and 
parents?].
We also ran RIPPLES with $p=4$, which reduced the overall number of 
events to 1174 (roughly 4-fold) but also approximately halved the number of 
events involving X lineages, including well-characterised events 
such as XBB (Table~\ref{tab:ripples_x_lineages}).

To compare these results with the sc2ts ARG, we first matched clades
by finding the most recent common ancestor (MRCA) of the samples
descending from each RIPPLES event. Overall, clade agreement was 
high with [FIXME, get exact numbers] 37XX nodes (90\%) in the sc2ts ARG where 
descendants match exactly. Of these nodes, 68\% had no recombinant
ancestry. Of the remainder, 595 could be matched with one of the 
855 recombination events in the sc2ts ARG
(Figure \ref{fig:ripples_sc2ts_events}).
Table~\ref{tab:ripples_x_lineages} shows a summary of all
recombination events associated with Pango X lineages detected by RIPPLES,
and their relationship with the sc2ts ARG.
Of the 16 Class I events (Pango X lineages unequivocally associated with
a recombination node in the sc2ts ARG) in Table~\ref{tab:pango_x_lineages},
8 have corresponding recombination events (4, at $p=4$).
Of the 5 Type II events
(recombination nodes with more complex assocation with X lineages),
two have close matches, with XBM matching exactly (including 2 descending BF.3 samples),
and XE/XH is similar (but without including the 2 XH samples in the clade).
Two events in the RIPPLES set correspond to likely false negatives 
in sc2ts: XB (Document~\ref{sec:dd_xb}) 
and XAJ (Document~\ref{sec:dd_xaj}).
See Document~\ref{sec:detailed_ripples} for more detailed analysis.

@@

@@

\subsection*{Prevalance of recombination in SARS-CoV-2}
[Rough first draft taken from earlier location. Sticking in a heading to
make it clear it's distinct from other paragraphs.]

This estimate for effective recombination in SARS-CoV-2 is much lower 
than previous studies have suggested.
% Bolotie: , we identified 225 anomalous SARS-CoV-2 genomes of likely
% recombinant origins out of the first 87,695 genomes to be released
In an early analysis, 225 potentially recombinant sequences out 
of 87,695 samples [WHAT DATE?]\cite{Varabyou2021rapid}.
Over the same time period, XX events are present in the sc2ts ARG,
representing a total of XX samples.
In the flagship analysis for the RIPPLES method~\cite{Turakhia2022}, 
589 unique recombination events in a tree of 1.6 million samples
up to May 2021 were identified after extensive quality control.
These nodes had a combined total of 43,104 descendants, suggesting
that approximately 2.7\% of genomes were of recombinant origin.
Over the same time period, the sc2ts ARG contains 14 recombinants
passing QC with a total of 172 descendants. The ARG contains a total
501,235 samples over this time period, giving an overall total fraction of
0.034\%. Thus, the sc2ts ARG estimates 42-fold fewer recombination
events and 79-fold descendant samples over the same time period.
The numbers are not directly comparable as the published RIPPLES 
estimate is based on GISAID data which is enriched for
artefacts~\cite{Hunt2024}, but given the observations in 
the previous section on running RIPPLES on the Viridian data
it seems likely that RIPPLES is overestimating the prevalence
of recombination in SARS-CoV-2.

@@

[NOTE: this is an earlier draft of main text which explains the table in
detail]
